% CODS-2025 Demo Track Paper
% AI Example Generator with Dynamic Learning Context Adaptation

\documentclass[sigconf]{acmart}

\usepackage{graphicx}
\usepackage{url}

% Conference information
\acmConference[CODS-COMAD'25]{ACM India Joint International Conference on Data Science and Management of Data}{December 17--20, 2025}{Pune, India}
\acmYear{2025}
\copyrightyear{2025}

% DOI and ISBN will be assigned by ACM
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}

% Permission statement
\setcopyright{acmlicensed}

\begin{document}

\title{ExaCraft: Dynamic Learning Context Adaptation for Personalized Educational Examples}

\author{Akaash Chatterjee}
\authornote{Both authors contributed equally to this research.}
\email{m24cse002@iitj.ac.in}
\author{Suman Kundu}
\authornotemark[1]
\email{suman@iitj.ac.in}
\affiliation{%
  \institution{Indian Institute of Technology Jodhpur}
  \city{Jodhpur}
  \state{Rajasthan}
  \country{India}
}

\begin{abstract}
Examples play a critical role in aiding the understanding of concepts. Learning effectiveness relies on relevant, relatable examples that connect with learners at a more personal level. Current educational AI systems often use static user profiles. These profiles can't adapt to changes in a learner's understanding, struggle patterns, or skill growth. To address this, we present ExaCraft, an AI system that revolutionizes personalized example generation through dynamic learning context adaptation. Integrating a Chrome extension, Python Flask API, and Google Gemini AI, ExaCraft combines user-configured profiles (location, education, profession, complexity preferences) with real-time behavioral analysis to deliver examples that are both culturally relevant and pedagogically adaptive. Its core innovation is dynamic adaptation based on five categories of behavioral learning context: struggle indicators, mastery patterns, topic progression history, session boundaries, and learning progression signals. The demonstration will showcase how examples evolve from basic conceptual illustrations to advanced technical implementations, based on real-time analysis of topic repetition, regeneration requests, and topic progression patterns. This work represents a significant advancement in personalized learning by combining static profile relevance with dynamic behavioral adaptation for enhanced comprehension.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003456.10003457.10003527.10003531</concept_id>
<concept_desc>Social and professional topics~Computing education</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003130</concept_id>
<concept_desc>Human-centered computing~Interactive systems and tools</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010257.10010293.10010311</concept_id>
<concept_desc>Computing methodologies~Natural language generation</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10002944.10011123.10011131</concept_id>
<concept_desc>General and reference~Empirical studies</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Social and professional topics~Computing education}
\ccsdesc[500]{Human-centered computing~Interactive systems and tools}
\ccsdesc[300]{Computing methodologies~Natural language generation}
\ccsdesc[300]{General and reference~Empirical studies}

\keywords{Educational Technology, Personalized Learning, Browser Extensions, AI-Generated Content, Learning Analytics, Context-Aware Systems, Dynamic Adaptation, Behavioral Learning Patterns}

\maketitle

\section{Introduction}

\subsection{Motivation}

Traditional education often uses generic examples. These do not connect well with learners' different backgrounds, experiences, and ways of thinking. This one-size-fits-all method for generating examples limits learning. Learners often struggle to connect with these examples.

The challenge becomes clearer when considering how each person learns differently. A data scientist learning quantum computing requires different examples than a physics student exploring the same concepts. Moreover, the same learner's example preferences evolve throughout their learning journey. A graduate software engineer might start with simple machine learning examples. As they learn more, they will need more advanced ones.

Current educational AI systems approach example generation through static personalisation by matching predetermined user categories to fixed example types. This approach sees personalised examples as a one-time match. It misses key signals about when learners need basic or advanced examples. When example generation does not adapt, learning experiences fall. This results in the examples' complexity, domain context, and presentation style to stay fixed. This static approach ignores changing comprehension levels and learner needs.

\subsection{Our System}

ExaCraft is an AI-powered example generation system that revolutionises personalised example generation through dynamic learning context adaptation. It is a Chrome browser extension that generates contextually-aware, personalised examples for any highlighted text, seamlessly integrating with users' natural web browsing workflows. The system is supported by a Python Flask API that combines Google Gemini AI with sophisticated learning analytics to deliver culturally relevant and pedagogically adaptive examples.

It implements a hybrid personalisation approach that combines user-configured static profiles with dynamic behavioural adaptation. Users define five key personalisation fields through the extension interface: name, location, education level, profession, and preferred complexity. This static foundation ensures cultural relevance and professional contextualisation. Building upon this foundation, ExaCraft's core innovation lies in generating personalised examples that evolve based on five categories of dynamic learning context derived from example generation interaction history: recent topic requests with timestamps, struggle indicators through example topic repetition patterns and regeneration button clicks, mastery signals via rapid progression through diverse example topics, session-level boundary tracking with cross-session example preference continuity, and behavioral signal integration across example generation episodes. This example-focused behavioural learning analytics approach enables ExaCraft to adapt example complexity and presentation style based on actual interaction patterns, while maintaining the cultural and professional relevance established through static profiling.

\subsection{Contributions}

This demonstration makes the following key contributions to educational AI research and practice:

\begin{itemize}
    \item \textbf{A hybrid personalisation framework} that combines user-configured static profiles (cultural, educational, professional context) with dynamic behavioural adaptation, tracking patterns across five contextual dimensions to generate personally relevant and pedagogically appropriate examples.
    
    \item \textbf{A seamless workflow integration} via a browser extension that provides zero-disruption example generation with integrated profile management, addressing critical adoption barriers in educational technology.
    
    \item \textbf{A behavioral analytics model for example adaptation} that uses multi-signal analysis (topic repetition, regeneration patterns) to detect struggle and mastery, dynamically adjusting example complexity while preserving static profile relevance.
    
    \item \textbf{A system for cross-session personalization continuity} that maintains both static preferences and dynamic adaptation patterns across multiple browsing sessions, enabling long-term, personalized learning progressions.
    
    \item \textbf{A fully deployed system for live demonstration} that showcases the practical application of hybrid static-dynamic example personalization in realistic educational scenarios.
\end{itemize}


\section{Related Work}

Educational personalization has been extensively studied in adaptive learning systems and intelligent tutoring systems. Traditional approaches focus on learner models based on knowledge states and skill assessments \cite{bayly2024personalized}. However, these systems typically rely on explicit assessments rather than implicit behavioral signals during natural learning activities.

Recent work in AI-driven educational content generation has explored using large language models for personalized content creation \cite{kasneci2023chatgpt}. While these systems demonstrate the potential for AI-generated educational materials, they primarily use static user profiles without incorporating real-time learning behavior adaptation.

Learning analytics research has investigated behavioral pattern recognition in online educational environments \cite{yagci2022educational}. Zhang et al. \cite{saqr2023temporal} explored session-based learning pattern recognition, focusing on temporal aspects of learning behaviors. However, existing approaches typically analyze learning patterns post-hoc for course improvement rather than providing real-time content adaptation.

Educational technology tools have increasingly focused on seamless integration with learners' natural workflows. Browser extensions and web-based learning tools have gained popularity for their non-disruptive approach to educational support. However, most existing tools provide static functionality without behavioral adaptation capabilities.

Struggle detection in educational systems has been primarily based on performance metrics such as quiz scores and completion times \cite{baker2019educational}. Our work extends this by introducing multi-signal behavioral analysis that includes interaction patterns, regeneration requests, and topic revisitation frequencies as indicators of learning difficulties.

Our system distinguishes itself by combining real-time behavioral learning analytics with seamless workflow integration, providing dynamic content adaptation based on implicit learning signals rather than explicit assessments or static user characteristics.

\section{System Architecture and Implementation}

Figure 1 illustrates ExaCraft's system architecture, showcasing the multi-flow interaction between the browser extension, API server, and dynamic learning context engine. The diagram shows four different operational flows: normal example generation (blue arrows), struggle detection and regeneration (red dashed arrows), session management (green dotted arrows), and data persistence (purple thin arrows). This comprehensive flow visualisation highlights how ExaCraft's behavioural learning analytics enable real-time adaptation of personalised examples based on user interaction patterns.

\subsection{Core Components}

ExaCraft consists of three integrated components working in concert to deliver personalized learning experiences:

\textbf{Browser Extension:} A Chrome extension provides smooth integration with web browsing workflows. Users highlight text and generate contextual examples via right-click menus. The extension features a user profile management interface where users specify five personalisation fields: name, location (city, country), education level (high school, undergraduate, graduate, professional), profession, and preferred complexity (simple, medium, advanced). Beyond profile management, it tracks real-time behaviour signals like regeneration patterns and topic repetition. These signals help adapt behaviour dynamically. As shown in Figure 1, the extension handles both regular generation requests (blue flow) and regeneration signals (red flow).

\textbf{API Server:} A Python Flask RESTful backend service orchestrates AI content generation with behavioural analytics. It receives requests from the extension, processes user profile data, tracks learning contexts, and integrates with Google Gemini AI through LangChain for refined prompt engineering. The API server processes these inputs and coordinates with the learning context engine to deliver examples adapted to user behaviour.

\textbf{Learning Context Engine:} This novel component tracks and analyses five categories of example generation behaviour in real time. Its core technical innovation is continuously analysing interaction patterns to build contextual models for real-time personalisation. For instance, the engine detects learning struggles primarily through example topic repetition frequency (three or more requests for the same topic) and explicit regeneration button clicks (red dashed flow). It identifies mastery through rapid progression across diverse topics (three or more different topics in recent history), which triggers an increase in the complexity of later examples. This engine maintains persistent storage through JSON files (purple flow) for cross-session continuity.

\subsection{Hybrid Static-Dynamic Personalisation Framework}

ExaCraft uses a two-layer system that combines user profiles with real-time learning behaviour analysis.

\textbf{Building the Foundation with User Profiles:} When users first set up ExaCraft, they fill out a straightforward profile in the browser extension. This includes their name, where they are from, their educational background (high school through professional), what they do for work, and how complex they want their examples to be. These details help create examples that actually make sense for each person. Take someone who is a graduate-level data scientist from Mumbai, they will get examples that reference familiar Indian contexts while diving into technical concepts that match their professional background.

\textbf{Learning from How People Actually Use the System:} Here is where things get interesting. The system watches how users interact with the examples it generates. Suppose someone repeatedly asks for examples on the same topic or clicks the regenerate button. In that case, ExaCraft recognises they might be struggling and automatically dials down the complexity, even if they initially said they wanted advanced content. Conversely, when users quickly move through different topics without getting stuck, the system picks up on their confidence. It starts offering more challenging examples while keeping the cultural and professional context they are familiar with.

This approach's beauty lies in balancing who learners are as individuals with how they are learning in the moment. Examples stay relevant to their background and interests, but they also adjust based on whether learners are having an easy day or wrestling with particularly tough concept.




\section{Live Demonstration Walkthrough}

The demonstration will showcase ExaCraft through three interactive scenarios that highlight key behavioral adaptation features. Attendees will observe the system in action through a browser session with pre-configured user profiles and learning contexts, experiencing the multi-flow architecture illustrated in Figure 1.

\subsection{Scenario 1: Adaptive Complexity Progression}

\textbf{Demo Setup:} The demonstration will begin with a simulated user (Profile: Graduate Student, Computer Science) browsing a machine learning tutorial webpage. Attendees will observe how personalized examples evolve through the following interaction sequence:

\begin{enumerate}
    \item User highlights "machine learning" → ExaCraft generates an intermediate-level personalized example leveraging the user's Computer Science academic background with programming analogies, algorithmic concepts, and cultural relevance (blue flow)
    \item User quickly progresses to "neural networks" and "deep learning" → ExaCraft detects mastery signals and adapts future examples with more sophisticated technical illustrations and implementation details
    \item User returns to "machine learning" and clicks regenerate button twice → ExaCraft records struggle signals (red dashed flow) and generates simplified personalized examples with foundational analogies and step-by-step conceptual breakdowns
    \item ExaCraft demonstrates how personalized examples transform from technical implementations to conceptual illustrations based on behavioral learning patterns captured through the system architecture
\end{enumerate}

\textbf{Key Observable Features:} Attendees will observe how personalized examples change in complexity based on behavioral signals, demonstrating how dynamic adaptation works alongside static profiles to adjust example difficulty while maintaining cultural and professional relevance.

\subsection{Scenario 2: Session-Aware Learning Continuity}

\textbf{Demo Setup:} The demonstration will showcase cross-session personalized example continuity through a multi-part scenario:

\begin{enumerate}
    \item User initiates a focused learning session on "data structures" → ExaCraft generates personalized examples using programming contexts familiar to the user (green dotted flow for session management)
    \item ExaCraft maintains personalized example preferences and complexity levels across topic progression through persistent context storage
    \item Session is ended and a new session is started after 5 minutes → ExaCraft recalls previous example topics and adapts based on established topic familiarity patterns (demonstrating cross-session continuity)
    \item User encounters difficulty with "binary trees" → ExaCraft recognizes previous familiarity with "arrays" topic and generates examples that leverage familiar array concepts to introduce tree structures, adapting complexity based on the established learning pattern
\end{enumerate}

\textbf{Key Observable Features:} The demonstration shows how personalized example preferences and topic familiarity patterns are maintained across sessions, enabling consistent yet adaptive personalized learning experiences that build upon previously explored concepts.

\subsection{Scenario 3: Multi-Signal Struggle Detection}

\textbf{Demo Setup:} The final scenario demonstrates sophisticated struggle detection for personalized example adaptation through multiple behavioral signals:

\begin{enumerate}
    \item User explores "quantum computing" with normal interaction patterns → ExaCraft generates standard complexity personalized examples appropriate to their profile (blue flow)
    \item User exhibits struggle signals: multiple regenerations, topic revisitation → ExaCraft automatically adapts to provide simplified personalized examples with familiar analogies and foundational conceptual illustrations (red dashed flow activation)
    \item Console output demonstrates how multiple behavioral signals (regeneration clicks + topic repetition) combine to trigger personalized example simplification while maintaining individual learning context through the integrated architecture flows
\end{enumerate}

\textbf{Key Observable Features:} Demonstration of behavioral signal detection for personalized example adaptation, showing how struggle signals trigger simplified examples while maintaining individual learning preferences.

Each scenario will include screenshots of the ExaCraft browser extension interface, example responses, and console output showing the learning context engine's behavior to demonstrate ExaCraft's example adaptation mechanisms.

\section{Implementation and Deployment}

ExaCraft is implemented using modern web technologies ensuring broad compatibility and ease of deployment. The Chrome extension uses Manifest V3 specifications for enhanced security and performance, while the Python Flask API leverages Google Gemini AI through LangChain for robust natural language processing capabilities.

All user data is stored locally in Chrome storage with optional server synchronization, ensuring privacy control while enabling cross-device learning continuity. ExaCraft includes comprehensive error handling and fallback mechanisms to ensure reliable operation even when AI services are temporarily unavailable.

The demonstration provides attendees with direct ExaCraft interaction through the live browser extension, enabling firsthand experience of how dynamic learning context adaptation enhances traditional static personalization approaches. ExaCraft is actively deployed with documented learning contexts for multiple test users, demonstrating practical applicability in real learning scenarios.

\section{Conclusion}

We have demonstrated ExaCraft, an AI-powered example generation system that addresses the limitations of static educational personalization through dynamic learning context adaptation. ExaCraft's novel behavioral learning analytics approach—tracking struggle indicators, mastery patterns, and session continuity—enables real-time content adaptation that responds to actual learning behaviors rather than predetermined user categories.

The live demonstration showcases practical applications of ExaCraft's approach through seamless browser integration and multi-signal behavioral analysis. By combining Google Gemini AI with sophisticated learning analytics, ExaCraft provides a foundation for next-generation educational AI applications that adapt to learners' evolving needs.

Future work could explore expanding ExaCraft's deployment to other platforms, such as Firefox extensions or integrated plugins for Learning Management Systems. We also aim to enhance the context engine to incorporate collaborative learning patterns and develop comprehensive evaluation frameworks for behavioral analytics in educational contexts.

\section{Demo Access Information}

Demo URL: Available upon request for conference review\\
Video Demo: Available upon request for conference review\\
Live Demo Requirements: Standard laptop with Chrome browser, internet connection\\
Source Code: Available upon request for conference review

\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

\end{document}